This is taken directly from the phosphoproteomics pipeline template .RMD. It will not work by itself. 

### Go Enrichment
GO enrichment checks if a subset of your data is enriched for a particular GO term vs your dataset as a whole. This idea isn't exclusive to GO terms, but it relies on expert annotation of everything in your dataset. We mostly trust GO annotations, but again, it's not perfect.

This current analysis is done at the PTM level, not adjusted...
```{r}
# only want to load these terms once. It takes a bit to run
if (!exists("global.go.terms")) {
  global.go.terms <- Load.GO.Terms.From.Bioconductor()
}

# This is all the proteins in your dataset.
dataset.all.proteins <- unique(plottableData$PTM.Model$Protein.No.Site)

# This is the subset of genes that you want to check for term enrichment
row.dendrogram <- row_dend(drawnHeatmap)

# get indices of clusters of interest based on dendrogram
cluster1.row.indices <- Dendrogram.Cluster.Get.Original.Indices(row.dendrogram[[2]][[1]][[1]])
cluster2.row.indices <- Dendrogram.Cluster.Get.Original.Indices(row.dendrogram[[2]][[1]][[1]])
cluster3.row.indices <- Dendrogram.Cluster.Get.Original.Indices(row.dendrogram[[2]][[2]][[1]])
cluster4.row.indices <- Dendrogram.Cluster.Get.Original.Indices(row.dendrogram[[2]][[2]][[2]])

# Get sitified proteins from cluster indices, then unsitify and unique them.
cluster1.proteins <- rownames(Sig.Changing.Site.Matrix)[cluster1.row.indices]
cluster1.proteins  <- unique(sapply(strsplit(as.character(cluster1.proteins ), "_"), function(x) x[1]))

cluster2.proteins <- rownames(Sig.Changing.Site.Matrix)[cluster2.row.indices]
cluster2.proteins  <- unique(sapply(strsplit(as.character(cluster2.proteins ), "_"), function(x) x[1]))

cluster3.proteins <- rownames(Sig.Changing.Site.Matrix)[cluster3.row.indices]
cluster3.proteins  <- unique(sapply(strsplit(as.character(cluster3.proteins ), "_"), function(x) x[1]))

cluster4.proteins <- rownames(Sig.Changing.Site.Matrix)[cluster4.row.indices]
cluster4.proteins  <- unique(sapply(strsplit(as.character(cluster4.proteins ), "_"), function(x) x[1]))

cluster.list <- list(Cluster_1 = cluster1.proteins, Cluster_2 = cluster2.proteins, Cluster_3 = cluster3.proteins, Cluster_4 = cluster4.proteins)

enrichment.list <- lapply (cluster.list, function(cluster.proteins){
                        setDT(as.data.table(clusterProfiler::enricher(gene = cluster.proteins, 
                                                                      universe = dataset.all.proteins, 
                                                                      TERM2GENE = global.go.terms,
                                                                      pAdjustMethod = "none",
                                                                      pvalueCutoff = 0.05)))
                       })

```

We're getting no enrichment for clusters 1 and 2. Let's summarize the top 10 enrichments for clusters 3 and 4
```{r}
#source(paste0(drb_utils,"gene_ontology_enrichment.R"))
cluster.enrichment.table <- rbindlist(enrichment.list[-c(1,2)], idcol = "Cluster")
perm.cluster.enrich.tab <- cluster.enrichment.table

enrichHeatmapBestPerGroup(cluster.enrichment.table, NULL, groupColumn = "Cluster", topN = 10)  # compare enrichment in the top 10 terms per group
                       
# GO.Enrichment.Multigroup(cluster.enrichment.table = cluster.enrichment.table,
#                          group.column = "Cluster", 
#                          p.value.column = "p.adjust", 
#                          p.value.threshold = 0.01,
#                          top.n = 10)
```