This code assumes the input data comes from the *MsStats::dataProcess* function, specifically the *\$ProteinLevelData*.

# Data Import

```{r}
# Normally this data will already be saved in memory
# Need to load it in here
proteinData <- fread("../example_data/example_data_dataProcess_output_protein.csv")

# if you're using the direct dataProcess output e.g. `dp_output$ProteinLevelData`
  ## make sure to cast it to a data table before going further
  ## otherwise your intensity values will be cast to character and things will break

# e.g. proteinData <- as.data.table(dp_output$ProteinLevelData)
```

# Create a matrix of complete protein cases

We need to do some data manipulation to get our dataset into the right shape for PCA.

-   Transform our data from long to wide format at the sample level, *i.e.* by *GROUP* and *SUBJECT*.
-   Subset our data to a set of completely profiled proteins

PCA does not work with missing values, so either you need to impute beforehand or filter as we do below

```{r}
# Go from long to wide format
# Row names are protein groups
# Columns are each sample's LogIntensity values for each protein
intensity.matrix <- as.matrix(dcast(as.data.table(experiment.dataProcessed$PTM$ProteinLevelData), Protein ~ SUBJECT, value.var = "LogIntensities"),
                           rownames = "Protein")


complete.data.matrix <- intensity.matrix[complete.cases(intensity.matrix),]
```

# Do PCA
```{r}
# Do PCA
pcaOut <- prcomp(t(complete.data.matrix))

# Reshape from matrix to data table
pcaDT <- as.data.table(pcaOut$x, keep.rownames=TRUE)

# Add column named `mainGroup`. 
# This value is derived from the rowname by splitting on '_' and grabbing the first element
pcaDT[, mainGroup := tstrsplit(rn, "_")[[1]]] #transpose & split

# Add column named `batch`. 
# This value is derived from the rowname by splitting on '_' and grabbing the first element
pcaDT[, batch := sapply(strsplit(rn, "_"), function(x) tail(x, 1))] #transpose & split

###
# You can add other metadata columns here too. This is just for my particular use case
###

pcaPercentVar <- round(100 * (pcaOut$sdev^2)/sum(pcaOut$sdev^2), 1)
```

# Visualize PCA Plots
A standard 2D representation of variance captured by PC1 and PC2
```{r}
# Visualize PC1 and PC2, color data points by batch value.
# You can also color data points by mainGroup or other factors that you scrape out above.
p <- ggplot (pcaDT, aes(x=PC1, y=PC2, color = batch )) + 
  geom_point(alpha=1.0, size=4) + 
  ggrepel::geom_text_repel(aes(label=rn), show.legend = FALSE, size = 3) +
  theme_bw() + 
  xlab (sprintf ("PC1, %.1f%%", pcaPercentVar[1])) + 
  ylab (sprintf ("PC2, %.1f%%", pcaPercentVar[2])) + 
  ggtitle (sprintf ("DOPA Samples - PCA using %d Phosphosites (log intensity)", nrow(complete.data.matrix))) 
p

# Optionally save the figure
ggsave(paste0("figures/", "Replicate_PCA.pdf"))
```

A 3D visualization done via Plotly. This allows you to also view PC3 at the same time.
```{r}
fig <- plot_ly(pcaDT, x = ~PC1, y = ~PC2, z = ~PC3, color = ~batch, colors = c('#636EFA','#EF553B','#00CC96') ) %>%
  add_markers(size = 12)


fig <- fig %>%
  layout(
    title = "3D PCA",
    scene = list(bgcolor = "#e5ecf6")
)

fig
```